### 2017-05-24 01-intro.tex
Merged cleanup from master

Starting from an existing implementation in Python/Numpy, we create an efficient
baseline version in C++ to serve as the baseline for the parallel versions.
This already improves performance by a factor of 1000 to 1500 compared to
Python. Then, our key contributions are:
\begin{itemize}
\item
our implementation is scaled up to run on GPUs.
We developed aggressive optimizations to solve the memory bandwidth issues for
GPUs. This implementation achieves speedup factors up to~86,
compared to the baseline.
%
Through careful analysis of the computation and data structures we show that
the algorithm's full state can be reduced by roughly 75\%. Compressing the
state significantly reduces the data intensity and allows for tackling larger
problems; % while maintaining all state in memory.

\item
our implementation is scaled out to overcome the memory bounds posed by GPUs or
single-machine systems in general, to solve the largest publicly available
community graphs. % This necessitated overcoming several challenges.
% The algorithm's state grows rapidly with larger graphs and number of latent
% communities.
Since the full state for the larger graphs does not fit in a
single machine's memory, it is partitioned across a cluster
of machines. Hence, the cluster nodes must read
remote memory hosted by their peers. Since the algorithm is memory-bound, we
substantially optimized remote memory accesses by implementing
a custom RDMA store.
On top of that, we pipelined the algorithm stages by fetching
data in advance over the network;
\begin{comment} Finally, the algorithm's computation
is effectively distributed across the cluster nodes and parallelized further
within each node by exploiting their multi-core CPUs.
\end{comment}

\item we argue that it is valid to use a CPU version with straightforward
intra-node thread parallelism for the distributed implementation, and
that it is pointless to combine the GPU work into it.

@@

The remainder of this paper is organized as follows.
Section~\ref{sec-background} provides an overview of the
algorithm and its theoretical foundation, with related work in
Section~\ref{sec-related}. Section~\ref{sec-design-sequential} describes how we
arrived at an efficient sequential version that serves as the baseline for the
parallel implementations. Section~\ref{sec-gpu} delves
into the parallelization optimizations for GPUs and multi-core CPUs.
Section~\ref{sec-distr} describes the design, implementation and evaluation of
the scalable distributed solution. Finally, Section~\ref{sec-conclusion}
provides concluding remarks.

>>>>

Starting from an existing implementation in Python/Numpy, we create an efficient
C++ version to serve as the baseline for the parallel versions.
This already improves performance by a factor of 1000 to 1500 compared to
Python. Then, our key contributions to the algorithm are:

\begin{itemize}
\item \textbf{scaling up to run on GPUs.}
The application is not a natural fit for GPUs, because it is memory-bound and
irregular in its data accesses. Nevertheless, we developed aggressive
memory allocation optimizations to solve the data bandwidth issues for GPUs.
This implementation achieves up to 86$\times$ speedup compared to the baseline.
%
We managed to reduce the algorithm's memory requirements by roughly 75\%, which
significantly reduces the data intensity and allows for tackling larger
problems; % while maintaining all state in memory.

\item \textbf{scaling out to overcome the memory bounds posed by GPUs} or,
generally, single-machine systems, to solve the largest publicly available
community graphs. % This necessitated overcoming several challenges.
% The algorithm's state grows rapidly with larger graphs and number of latent
% communities.
Since the full state for the larger graphs does not fit in a
single machine's memory, it is partitioned across a cluster
of machines. Hence, the cluster nodes must read
remote memory hosted by their peers. Since the algorithm is memory-bound, we
substantially optimized remote memory accesses by implementing
a custom RDMA store.
On top of that, we pipelined the algorithm stages by fetching
data in advance over the network;
\begin{comment} Finally, the algorithm's computation
is effectively distributed across the cluster nodes and parallelized further
within each node by exploiting their multi-core CPUs.
\end{comment}

\item \textbf{the use of pipelining} allows to use straightforward intra-node
thread parallelism in the distributed implementation, and it is pointless
to combine the GPU work into it. The cause is that the computation can be
fully overlapped with fulfilling the data requests for the next iteration.

@@

The remainder of this paper is organized as follows.
Section~\ref{sec-background} provides an overview of the
algorithm and its theoretical foundation, with related work in
Section~\ref{sec-related}. Section~\ref{sec-design-sequential} describes how we
arrived at an efficient sequential version that serves as the baseline for the
parallel implementations. Section~\ref{sec-gpu} delves
into the parallelization optimizations for GPUs.
Section~\ref{sec-distr} describes the design, implementation and evaluation of
the scalable distributed solution. Finally, Section~\ref{sec-conclusion}
provides concluding remarks.
